# Clustering

## Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(scran)
library(scater)
library(ggspavis)
library(patchwork)
library(SpatialExperiment)
# load data from previous section
spe <- readRDS("img-spe_qc.rds")
# basic theme for spatial plots
theme_xy <- list(
    coord_equal(expand=FALSE), 
    theme_void(), theme(
        plot.margin=margin(l=5),
        legend.key=element_blank(),
        legend.key.size=unit(0.5, "lines"),
        panel.background=element_rect(fill="black", color="white")))
```

## Normalization

Library size-based normalization as is typical for scRNA-seq data has been shown to be problematic for ST data, especially so for targeted panels underlying current commercial imaging-based ST platforms [@Atta2024,@Bhuva2024].
For lack of a better approach, we here use standard log-library size normalization.
We caution readers, however, to keep an eye out in the literature for attempts to provide a better strategy.

```{r calc-norm}
spe <- logNormCounts(spe)
```

## Analysis

### Other methods

- `r BiocStyle::Biocpkg("SpatialPCA")`

### Spatially aware

`r BiocStyle::Biocpkg("BANKSY")` [@Singhal2024-BANKSY] essentially computes PCs that are weighted by embeds cells in a product space of their own and the local neighborhood transcriptome, representing cell state and microenvironment.
A key parameter for this method is $\lambda\in[0,1]$ (argument `lambda` in `runBanksyPCA()`), which controls the amount of spatial information to incorporate; specifically, **when $\lambda=0$, `BANKSY` reduces to non-spatial clustering**.

```{r clust-sp, message=FALSE, cache=TRUE}
library(Banksy)
spe <- computeBanksy(spe, assay_name="logcounts")
spe <- runBanksyPCA(spe, npcs=20, lambda=0.2)
spe <- clusterBanksy(spe, algo="louvain", resolution=0.5)
k <- grep("^clust", names(colData(spe)))
names(colData(spe))[k] <- "k_sp"
table(spe$k_sp)
```

### Transcription-only

As a baseline, we will use a standard scRNA-seq approach, namely, (spatially unaware) graph-based clustering based on a shared nearest neighbor (SNN) graph and the Louvain algorithm for community detection.

```{r clust-tx, message=FALSE, cache=TRUE}
library(igraph)
spe <- runPCA(spe, ncomponents=20) 
g <- buildSNNGraph(spe, use.dimred="PCA", k=30)
k <- cluster_louvain(g, resolution=0.5)$membership
table(spe$k_tx <- factor(k, seq_along(unique(k))))
```

### Comparison

To not confuse different types of PCs, we rename `reducedDims` to end in `_sp` and `_tx` for spatially aware and unaware results, respectively.
For custom visualizations, we then collect cell metadata and spatial coordinates:

```{r prep-data}
reducedDimNames(spe) <- c("PCA_sp", "PCA_tx")
df <- data.frame(colData(spe), spatialCoords(spe))
```

Visualizing the tissue with cells colored by their cluster assignment, we see pronounced spatial organization of subpopulations in both approaches:

```{r plot-xy-k-joint, fig.width=8, fig.height=3}
#| collapse: true
nk <- max(
    length(unique(df$k_sp)), 
    length(unique(df$k_tx)))
pal <- hcl.colors(nk, "Spectral")
ggplot(df, 
    aes(x_centroid, y_centroid, col=k_sp)) +
    ggtitle(expression("Banksy clustering ("*lambda~"0.2)")) +
ggplot(df, 
    aes(x_centroid, y_centroid, col=k_tx)) +
    ggtitle("SNN graph-based Louvain clustering") +
plot_layout(nrow=1) &
    geom_point(shape=16, stroke=0) &
    scale_color_manual(values=pal) & theme_xy &
    guides(col=guide_legend(override.aes=list(size=2)))
```

Especially in large tissues, the above type of plot makes it difficult to spot rare subpopulations, and might cause cells to overlap in regions with high cellular density.
This can be misleading, as we will tend to see only highly abundant subpopulations, or the cells plotted last and on top (i.e., later columns in the object).

To better distinguish between different subpopulations, we can instead generate separate spatial plots with one subpopulation highlighted at a time:

```{r plot-xy-k-split, collapse=TRUE, fig.width=12, fig.height=4}
ps <- lapply(unique(df$k_tx), \(.) 
    ggplot(df, aes(x_centroid, y_centroid, col=(k_tx == .))) + 
    scale_color_manual(values=c("navy", "cyan")) + ggtitle(.))
wrap_plots(ps, nrow=2) & 
    coord_equal(expand=FALSE) &
    geom_point(shape=16, stroke=0, size=0.2) &
    theme_void() & theme(
        legend.position="none",
        plot.title=element_text(hjust=0.5),
        panel.background=element_rect(fill="black", color="white"))
```

For details on identifying genes that are differentially expressed (DE) between groups of cells, we refer readers to [OSCA](https://bioconductor.org/books/3.19/OSCA.basic/marker-detection.html); a standard approach is given below, visualizing the average expression of exemplary markers across clusters:

```{r mgs}
#| code-fold: true
mgs <- findMarkers(spe, groups=spe$k_tx, direction="up")
top <- lapply(mgs, \(.) rownames(.)[.$Top <= 1])
top <- unique(unlist(top))
pbs <- aggregateAcrossCells(spe, 
    ids=spe$k_tx, subset.row=top, 
    use.assay.type="logcounts", statistics="mean")
plotHeatmap(pbs, 
    colour_columns_by="k_tx", fontsize=9,
    features=top, center=TRUE, scale=TRUE,
    cluster_cols=FALSE, cluster_rows=FALSE)
```

```{r plot-xy-expr, fig.width=4, fig.height=3}
#| collapse: TRUE
es <- as.matrix(t(logcounts(spe)[top, ]))
fd <- data.frame(df, es, check.names=FALSE)
ggplot(fd, aes(x_centroid, y_centroid, col=KRT14)) +
    coord_equal(expand=FALSE) +
    geom_point(shape=16, stroke=0, size=0.2) +
    scale_color_gradientn(colors=c("navy", "red", "gold", "ivory")) +
    theme_xy + theme(legend.key.height=unit(1, "lines"))
```

## Transfer

`r BiocStyle::Biocpkg("InSituType")` [@Danaher2022-InSituType] has been developed for CosMx data, specifically, but should be applicable to other types of imaging-based ST data.
It relies on a probabilistic model, can incorporate non-transcriptional information (e.g., immunofluorescence, cell area and aspect ratio), and can be run in an un-, semi- or fully supervised setting (based on bulk references profiles for the latter two).

Reference profiles may be acquired from custom reference data, e.g., from matching scRNA-seq data, or snPATHO-seq data on adjacent sections.
Some profiles are available through `r BiocStyle::Biocpkg("SpatialDecon")` [@Danaher2022-SpatialDecon].

```{r SpatialDecon, eval=FALSE}
library(SpatialDecon)
ref <- download_profile_matrix(
    species="Human", 
    age_group="Adult", 
    matrixname="ImmuneTumor_safeTME")
ref[1:5, 1:5]
colnames(ref)
```

```{r InSituType, message=FALSE, eval=FALSE}
library(InSituType)
library(STexampleData)
spe <- CosMx_lungCancer()
# cohorting based on IF data
j <- names(cd <- colData(spe))
i <- grep("^Mean", j, value=TRUE)
i <- setdiff(i, "Mean.G")
i <- c("Area", "AspectRatio", i)
coh <- fastCohorting(as.matrix(cd[i]))
# estimate per-cell background
rna <- spe$nCount_RNA
neg <- spe$nCount_negprobes/20
bkg <- rna*mean(neg)/mean(rna) 
# updating reference profiles 
# to account for platform effects
ref <- updateReferenceProfiles(
    reference_profiles=pbs, reference_sds=NULL, 
    counts=mtx, neg=neg, bg=bkg)$updated_profiles
# run 'InSituType' clustering
# in fully supervised mode
ist <- insitutype(mtx, 
    reference_profiles=ref, 
    update_reference_profiles=FALSE,
    neg=neg, cohort=coh, n_clusts=0)
table(ist$clust)
```

## Reduction

Exploratory data analysis (EDA) is crucial at every step; first and foremost after clustering and before performing downstream analyses towards biological insights.
A useful visualization is to color cells by their PCs in physical space. 
This will help highlight key drivers of transcriptional variability, e.g., between major biological compartments such as epithelia, immune and stromal cells.
Exemplary plots of PCs 1 and 2 are rendered below:

```{r plot-xy-pcs, collapse=TRUE, fig.width=8, fig.height=3}
#| code-fold: true
df <- data.frame(df, reducedDim(spe, "PCA_tx"))
ggplot(arrange(df, abs(PC1)), aes(x_centroid, y_centroid, col=PC1)) +
ggplot(arrange(df, abs(PC2)), aes(x_centroid, y_centroid, col=PC2)) +
    plot_layout(nrow=1) &
    coord_equal(expand=FALSE) &
    geom_point(shape=16, stroke=0, size=0.2) &
    scale_color_gradientn(colors=c("cyan", "navy", "magenta")) &
    theme_void() & theme(
        legend.position="bottom",
        legend.key.size=unit(0.5, "lines"),
        panel.background=element_rect(fill="black", color="white"))
```

::: {.callout-note title="PCs as RGB" collapse="true"}

#### PCs as RGB

A neat trick in this context is to rescale PC 1-3 values between 0 and 1, and interpret them as RGB values for visualization:

```{r pcs-to-rgb}
pcs <- reducedDim(spe)[, seq_len(3)]
pcs <- sweep(pcs, 2, colMins(pcs), `-`)
pcs <- sweep(pcs, 2, colMaxs(pcs), `/`)
df$rgb <- apply(pcs, 1, \(.) do.call(rgb, as.list(as.numeric(.))))
```

E.g., we see that colors corresponding to opposing PC values are clearly separated in space, indicating distinct drivers of transcriptional variability.
In some regions, we see color mixtures, indicating cells of ambiguous transcriptional profiles (with respect to what is being captured by the 3 PCs considered here).

```{r plot-xy-rgb, fig.width=4, fig.height=3}
ggplot(df, aes(x_centroid, y_centroid, col=rgb)) +
    geom_point(shape=16, stroke=0, size=0.2) +
    scale_color_identity() + theme_xy
```

:::

::: {.callout-note title="PC regession" collapse="true"}

#### PC regession

For any single-cell analysis where downstream tasks rely on PCs, it is useful to perform linear regression of (continuous or categorical) covariates of interest onto PCs.
This quantifies the variance explained by the covariate and can help assess the extend of unwanted variation (due to, e.g., cell area) as opposed to subpopulations driving transcriptional differences.
Here, we regress total counts, cell area, and cluster assignments against PCs:

```{r calc-pcr}
pcs <- reducedDim(spe, "PCA_tx")
ids <- c("total_counts", "cell_area", "k_tx", "k_sp")
pcr <- lapply(ids, \(id) {
    fit <- summary(lm(pcs ~ spe[[id]]))
    r2 <- sapply(fit, \(.) .$adj.r.squared)
    data.frame(id, pc=seq_along(r2), r2)
}) |> 
    do.call(what=rbind) |>
    mutate(id=factor(id, ids))
```

Here, `k_tx` (transcription-only) and `k_sp` (spatially aware) clusterings perform similar in terms of capturing (spatially unaware) PCs.
Things might look differently had we set a higher value for $\lambda$ when running `Banksy`; conversely, we would expect results to converge for $\lambda=0$.

```{r plot-pcr, warning=FALSE, fig.width=5, fig.height=2.5}
#| code-fold: true
pal <- c("magenta", "gold", "cyan", "blue")
ggplot(pcr, aes(pc, r2, col=id)) +
    geom_line(show.legend=FALSE) +
    geom_point() +
    coord_cartesian(xlim=c(1, 20)) +
    guides(col=guide_legend(override.aes=list(size=2))) +
    scale_x_continuous(breaks=c(1, seq(5, 20, 5))) +
    scale_y_continuous(limits=c(0, 1), breaks=seq(0, 1, 0.2)) +
    labs(x="principal component", y="coeff. of determination") +
    scale_color_manual("predictor", values=pal) +
    theme_minimal() + theme(
        panel.grid.minor=element_blank(),
        legend.key.size=unit(0.5, "lines"))
```

:::

```{r save-data}
# default to transcription-based clustering
spe$k <- spe$k_tx; saveRDS(spe, "img-spe_cl.rds")
```

## References {.unnumbered}
